---
title: "ticphasetype"
output: 
  pdf_document
vignette: >
  %\VignetteIndexEntry{ticphasetype}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ticphasetype)
```

#$T_{MRCA} and T_{Total}$

In an evolutionary tree the time until two sequences coalesce $T_i$ can be measured in number of generations $R_i$ divided by the population size $N$, this is, $T_i=R_i/N$. $T_i$ can easily be proven to approximate to an exponential distribution with rate $\binom{i}{2}$.

In order to understand the evolutionary history of sequences two additional quantities can be defined --namely the time until the most recent common ancestor $T_{MRCA}$ and the total tree length $T_{Total}$. $T_{MRCA}$ will simply be the sum of all times until two sequences coalesce, in other words $T_{MRCA}=T_n+T_{n-1}+...+T_2$, where $T_i\sim\binom{i}{2}$. $T_{Total}$, on the other hand, takes into account the length of all possible branches, so $T_{Total}=nT_n+(n-1)T_{n-1}+...+2T_2$ and, thus, $iT_i\sim \text{exp}(\binom{i-1}{2})$.

The mean and variance of these two quantities can be derived relatively easy. Defining their distribution, however, has proven to be more challenging since both $T_{MRCA}$ and $T_{Total}$ are sums of independent exponentially distributed variables with different rates. Their distribution can be computed as a series of convolutions, but their formulation, application and interpretation might be challenging for the average population geneticist.

Instead, we can think of the sum of exponential distributions as a continuous-time Markov chain, where coalescent events are represented as Markov jumps with rate $T_i$ for $T_{MRCA}$ and $iT_i$ for $T_{Total}$. The Markov chain will end with an absorbing state, which in both cases will be the MRCA. 

The Markov chain can be represented using phase-type theory, where the jump rates are defined with a sub-intensity matrix $T$ and the initial distribution will be defined as a row vector $\pi$. If we define $\tau$ as the smallest time (or length) for which we are in the absorbing state, then $\tau\sim PH(\pi,T)$. This continuous phase-type distribution has well-documented and easy-to-implement formulas for the expectation, the variance, the survival function, the distribution function and the density function. Moreover, since both $\pi$ and $T$ can easily be speficied, we can easily represent evolutionary histories that do not follow the standard coalescent model and still use the same phase-type formulas. 

`ticphasetype` contains an efficient implementation of continuous phase-type distributions. It has a user-friendly interface for creating phase-type representations of $T_{MRCA}$ and $T_{Total}$ under the standard coalescent model, but it also allows the user to specify their own sub-intensity matrix and initial probabilities for a more flexible implementation. 

A `phase_type` class representing a continuous phase-type distribution for $T_{MRCA}$ can be generated using `phase_type()`. For example, when `n`=5:

```{r}
T_MRCA_ph <- phase_type(type = 'T_MRCA', n = 5)
```

There are a number of methods associated with the `phase_type`, such as:

```{r}
mean(T_MRCA_ph)
```

```{r}
var(T_MRCA_ph)
```

```{r}
summary(T_MRCA_ph)
```

$T_{Total}$ can also be represented using the `phase_type` class:

```{r}
T_Total_ph <- phase_type(type = 'T_Total', n = 5)
```

```{r}
mean(T_Total_ph)
```

```{r}
var(T_Total_ph)
```

```{r}
summary(T_Total_ph)
```

Moreover, the height of an evolutionary tree can also be represented by a user-defined sub-intensity matrix. This is specially useful if the model does not follow Kingman’s n–coalescent, but other coalescent models such as the psi-coalescent or the beta-coalescent. As an example, an arbitrary sub-intensity matrix can be generated:

```{r}
n=5
subint_arbit <- matrix(0, nrow = n-1, ncol = n-1)

for (i in 1:(n-1)) {
  for (j in 1:(n-1)) {
    if (j>i) {
      subint_arbit[i,j] <- runif(1, 1, 10)
    }
  }
  subint_arbit[i,i] <- -sum(subint_arbit[i,])
}

subint_arbit[n-1,n-1] <- -1
```

```{r}
subint_arbit
```


This matrix can be supplied to the `phase_type()` generator function, together with optional initial probabilities:

```{r}
T_arbit_ph <- phase_type(subint_mat = subint_arbit)
```

If the initial probabilities are not supplied (as is the case above), then `phase_type()` automatically generates a vector of initial probabilities as $\pi=(1,0,...,0)$ and raises a warning message. 

We can apply the methods `phase_type` class methods for this new user-tailored phase-type distribution:

```{r}
mean(T_arbit_ph)
```

```{r}
var(T_arbit_ph)
```

```{r}
summary(T_arbit_ph)
```





















